D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-01-27 ootavacuous.tex
January 26th reorganization, part 2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Roughly speaking and all other things being equal, \emph{there is semantic
dependency from a given load to a given store when a change in the value
loaded that can change the value stored.}

@@

There is a semantic dependency from \co{y} to \co{x} if and only if the
value of \co{z} is non-zero.
Otherwise, in the absence of observable behavior (such as volatile
accesses), the implementation is within its rights to discard this
statement and act as if later accesses to \co{x} were instead the
constant zero, at least up to the next assignment to \co{x}.

Worse yet, suppose that both \co{y} and \co{z} are zero.
Then there is no semantic dependency from either \co{y} or \co{z} to
\co{x} unless both \co{y} and \co{z} change.
Alternatively, when both \co{y} and \co{z} are zero, there is no semantic
dependency from either \co{y} or \co{z} to \co{x}, but rather a semantic
dependency from the combination of \co{y} \emph{and} \co{z} to \co{x}.
This means that precise definitions of sdep cannot always rely on
single-variable perturbations.
See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
for examples and additional discussion.

Consider also:

@@

Also, consider this:

@@

\subsection{Difficulties in Rigorously Defining ``Semantic Dependency''}

But beware of this example:

@@

Because the stores executed on each branch of that "if" statement store
identical values to identical addresses, one could equally well regard
the two statements as performing two different stores or as performing
(for all intents and purposes) a single store.
Although software projects relying on ordering from load-to-store control
dependencies and using a variety of compilers should take the single-store
viewpoint~\cite{Howells2009membartxt},
the fact remains that these two viewpoints lead to differing conclusions
about whether there is a semantic dependency.

@@

These examples demonstrate a key point: Although in some cases sdep can
be a strict function of the source code, \emph{in general, sdep must be
defined on a per-execution basis}.% \footnote{
% 	Thanks to Alan Stern for the examples in this section and
% 	in the preceding one.}

\subsubsection{Semantic Dependencies Depend on Context}
\label{sec:Semantic Dependencies Depend on Context}

@@

\emph{Any analysis of semantic dependencies must either be fully informed
of the enclosing context, or much make conservative assumptions about
that context.}

\subsubsection{Semantic Dependencies Can Be Nondeterministic}
\label{sec:Semantic Dependencies Can Be Nondeterministic}

@@

\subsubsection{Semantic Dependencies Can Be Complex}
\label{sec:Semantic Dependencies Can Be Complex}

@@

Here, the presence or absence of a semantic dependency from \co{y} to
\co{x} depends not only on the value of \co{z}, but also the code in a
function that, although single threaded, is horribly complex.

@@

The key point here is that diferent values might cause almost completely
disjoint code to be executed.

@@

A similar example could be constructed from Fermat's Last Theorem
(FLT),\footnote{
\url{https://en.wikipedia.org/wiki/Fermat\%27s_Last_Theorem}}
which was first stated in the year 1637 and finally proved more than
300~years later, in 1994.
Therefore, it was only after 1994 that code relying on FLT was known
not to have a semantic dependency.

@@

\emph{Because of unsolved mathematical problems, precise definitions of
``semantic dependency'' have limitations.
Therefore, C++ implementations must assume that a given dependency is
a semantic dependency unless and until it can prove otherwise.}

\subsubsection{Implementers and Users Influence Semantic-Dependencies Definition}
\label{sec:Implementers and Users Influence Semantic-Dependencies Definition}

@@

The standard does not specify whether or not type \co{char} is signed,
leaving that choice to the implementations, which follow the lead
of the architects of a given CPU family.
Therefore, the code above has a a semantic dependency when run on systems
choosing signed \co{char}, but not when run on systems choosing unsigned
\co{char}.
So score one for the implementers, albeit by conscious choice on the
part of the standards committee.

Except that GCC provides the \co{-funsigned-char} command-line
argument that causes this implementation to treat variables of
type \co{char} as unsigned, regardless of the architects' wishes.
Given this command-line argument, there is always a semantic dependency
from the \co{foo()} function's argument to its return value.
So score one for at least some implementations' users.

@@

\emph{A definition of sdep drawn strictly from the standard will be
at best an approximation to a definition that is useful in practice.}

@@

>>>>

Roughly speaking and all other things being equal, \emph{there is semantic
dependency from a given load to a given store when a change in the value
loaded can change the value stored.}

@@

As long as \co{z} is zero, and all else being equal, changes in the
value of \co{y} will not cause changes to the value of \co{x}.
In that case, the implementation is within its rights to discard this
statement and act as if later accesses to \co{x} were instead the
constant zero, at least up to the next assignment to \co{x}.

\subsection{Semantic Dependencies Can Be Many To One}
\label{sec:Semantic Dependencies Can Be Many To One}

Suppose that both \co{y} and \co{z} are zero in the previous example.
Then changes to either but not both of \co{y} or \co{z} (again, all
else being equal) will not cause changes to the value of \co{x}.
In other words, there is no semantic dependency from either \co{y}
or \co{z} to \co{x} unless both \co{y} and \co{z} change.
Alternatively, when both \co{y} and \co{z} are zero, there is no semantic
dependency from either \co{y} or \co{z} to \co{x}, but rather a semantic
dependency from the combination of \co{y} \emph{and} \co{z} to \co{x}.
This means that precise definitions of sdep cannot always rely on
single-variable perturbations.
See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
for examples and additional discussion.

Roughly speaking and all other things being equal, \emph{there is semantic
dependency from a given set of loads to a given store when a change in
the values loaded can change the value stored.}

This definition allows the definition of an OOTA cycle to be expanded
from \co{acyclic(sdep ; rfe)} to:

\begin{quote}
	There are stores
	$W_i (i = 0, ..., m)$,
	where $W_i$ semantically depends on loads
	$R_{i,j} (j = 0, ..., n_i)$,
	such that each $R_{i,j}$ reads from one of the $W_k$
	stores in a different thread.
\end{quote}

\subsection{Semantic Dependencies Affected by Optimizations}
\label{sec:Semantic Dependencies Affected by Optimizations}

@@

\subsection{Semantic Dependencies Affected by \co{if} Statements}
\label{sec:Semantic Dependencies Affected by if Statements}

Consider the following simple \co{if} statement:

@@

\subsection{What Exactly is an Execution?}
\label{sec:What Exactly is an Execution?}

@@@ Roadmap

\subsubsection{Executions Can Be Merged}
\label{sec:Executions Can Be Merged}

Consider this example, which appears to be an \co{if} statement:

@@

Because the stores executed on each branch of that \co{if} statement store
identical values to identical addresses, one could equally well regard
the two statements as performing two different stores or as performing
(for all intents and purposes) a single store.
Although software projects relying on ordering from load-to-store control
dependencies and using a variety of compilers should take the single-store
viewpoint~\cite{Howells2009membartxt},
the fact remains that these two viewpoints lead to differing conclusions
about whether this code represents one execution or two.
It is the compiler's choice.

@@

These examples demonstrate a key point: 
\emph{In general, executions do not necessarily map one-to-one onto
paths through the source code}.% \footnote{
% 	Thanks to Alan Stern for the examples in this section and
% 	in the preceding one.}

\subsubsection{Executions Depend on Context}
\label{sec:Executions Depend on Context}

@@

\emph{Any analysis of a given execution must either be fully informed
of the enclosing context, or much make conservative assumptions about
that context.}

\subsubsection{Executions Can Be Nondeterministic}
\label{sec:Executions Can Be Nondeterministic}

@@

\subsubsection{Executions Can Be Complex}
\label{sec:Executions Can Be Complex}

@@

Here, the presence or absence of a semantic dependency from \co{y} to
\co{x} depends not only on the value of \co{z}, but also and execution
that passes through the code in a function that, although single threaded,
is horribly complex.

@@

The key point here is that diferent values might result in almost
completely disjoint executions.

@@

A similar example could be constructed from Fermat's Last Theorem
(FLT),\footnote{
\url{https://en.wikipedia.org/wiki/Fermat\%27s_Last_Theorem}}
which was first stated in the year 1637 and finally proved more than
300~years later, in 1994.
Therefore, it was only after 1994 that certain types of executions
relying on FLT were known not to have a semantic dependency.

@@

\emph{Because of unsolved mathematical problems, precise definitions of
``execution'' have limitations.
Therefore, C++ implementations must assume that a dependency passing
through a given execution is a semantic dependency unless and until it
can prove otherwise.}

\subsubsection{Implementers and Users Influence the Definition of Execution}
\label{sec:Implementers and Users Influence the Definition of Execution}

@@

The standard does not specify whether or not type \co{char} is signed,
leaving that choice to the implementations, which follow the lead
of the architects of a given CPU family.
Therefore, the code above has a semantic dependency in executions running
on systems choosing signed \co{char}, but not those running on systems
choosing unsigned \co{char}.
So score one for the implementers, albeit by conscious choice on the
part of the standards committee.

Except that GCC provides the \co{-funsigned-char} command-line
argument that causes this implementation to treat variables of
type \co{char} as unsigned, regardless of the architects' wishes.
Given this command-line argument, there is always a semantic dependency
in the execution running from the \co{foo()} function's argument to its
return value.
So score one for at least some implementations' users.

@@

\emph{A definition of ``execution'' drawn strictly from the standard will
be at best an approximation to a definition that is useful in practice.}

\subsubsection{So What Exactly is an Execution, Anyway?}
\label{sec:So What Exactly is an Execution, Anyway?}

Given all of these complications in C++ executions, and given that
semantic dependencies must be defined in terms of executions, it makes
sense to lower our sights from full generality.
We therefore restrict ourselves to a specific context:
traditional compilers.
Not only that, we will take the notion of semantic dependence (and thus
OOTA) to be per-compiler rather than absolute.

For the special case of non-volatile atomic operations, we will further
restrict ourselves to compilers that do single-threaded analysis, with
the possible addition of specific constraints deduced from multi-threaded
analysis.
For volatile atomic operations, compilers are prohibited from inventing
loads, and we will restrict ourselves to compilers that extend that
prohibition to non-volatile atomic loads.

@@@ LTO?

These restrictions permit us to bring to bear the real-world
constraints introduced by the next section.

@@@  Need to add material on inventing loads, which is currently
in an appendix.

@@

\subsection{But what if you want source-code dependencies?}

\subsubsection{@@@ Semantic Dependencies Depend on Context}
\label{sec:Semantic Dependencies Depend on Context}

\subsubsection{@@@ Semantic Dependencies Can Be Nondeterministic}
\label{sec:Semantic Dependencies Can Be Nondeterministic}

@@@ What do we do with this one???  Not sure we need it.

@@@ Not sure we need this one.

\subsubsection{@@@ Semantic Dependencies Can Be Complex}
\label{sec:Semantic Dependencies Can Be Complex}

\subsubsection{@@@ Implementers and Users Influence Semantic-Dependencies Definition}
\label{sec:Implementers and Users Influence Semantic-Dependencies Definition}
